A Classical Analog of Quantum Interaction-Free Measurement

### 2015-07-17 ms/ifm.tex
editing

\title{Quantum interaction-free measurement and a classical analog}

@@

\begin{abstract}
Interaction-free measurement refers to quantum phenomena such as the
presence of a photon absorber being inferred without it absorbing any
photons.  We review the physics and its realization in interferometric
experiments.  We then consider a simple classical system: two coupled
oscillators, one of which is damped; the energy dissipation can be
made arbitrarily small by making the damping large enough.  The
phenomenon can be considered analogous to interaction-free
measurement.
\end{abstract}

@@

Many quantum paradoxes have been explored theoretically and
experimentally in recent years.  Most involve entangled states: the
famous EPR paradox is an old example, and quantum computing is a more
recent development\cite{2007qucosc.book}.  Entanglement is considered
an intrinsically quantum phenonemn classical analog.

There are also counter-intuitive quantum phenonema not involving
entanglement.  Imagine a photon, initially in a ``vertical'' polarized
state.  It is sent through a Faraday rotator, and then through a
vertically polarizing sheet.  The former rotates the polarization
plane towards ``horizontal'' by small angle, say $\pi/(2N)$, where $N$
is large.  The latter effectively measures the polarization state in
the vertical/horizontal basis; if the result comes out horizontal, the
photon is absorbed, otherwise it goes through in a vertical
eigenstate.  Since $N$ is large, the chances of being absorbed are
small.  Most likely, the photon will get slightly rotated and then
reset to vertical.  If the photon is re-routed back to the apparatus
$N$ times, it will still most likely emerge vertically polarized.  The
rotator and the polarizer appear to cancel each other out.  The
general phenomenon is known as the quantum Zeno
effect,\cite{PhysRevA.41.2295} but it can also be described in
everyday language, in terms of waking and sleeping
states.\cite{orzel2009teach}

What happens if we now remove the polarizer?  Clearly, the $N$ Faraday
rotations will take the photon to a horizontal state.  Thus, the net
effect of the polarizer is (most likely) to rotate a photon from
horizontal to vertical.  Recall that the polarizer is really just an
absorber of horizontal photons.  Yet it can, within a clever setup, be
made to rotate photons instead of absorbing them.  This is the essence
of interaction-free measurement.  Since the initial suggestion
(essentially the $N=1$ case)\cite{1993FoPh...23..987E} and
generalization to the now-standard form, together with the first
experiments\cite{PhysRevLett.74.4763} interaction-free measurement has
been the subject of many
demonstrations\cite{voorthuysen:1504,2006JPhB...39.3177N} and
reviews,\cite{deweerd:272} and applications aiming towards medical
imaging are starting to appear\cite{PhysRevA.80.040902}.  (An older
usage of similar
terminalogy\cite{springerlink:10.1007/BF01327019,dicke:925} refers to
a situation where a non-detection enables an inference.)

@@

\section{Quantum formulation}

@@

With these concrete descriptions in mind, let us formalize the process
of interaction-free measurement.  We write the two eigenstates in the
notation of cubits
\begin{equation}
\ket0 \equiv \left( \begin{array}{c} 1 \\ 0 \end{array} \right)  \qquad
\ket1 \equiv \left( \begin{array}{c} 0 \\ 1 \end{array} \right)
\end{equation}
and
\begin{equation}
\X \equiv \left( \begin{array}{cc} 0 & 1 \\ 1 & 0 \end{array} \right)
\end{equation}
denotes a swap operator.  A beam splitter is given by the unitary operator
\begin{equation}
\exp\left[i\left({\textstyle\frac\pi2-\epsilon}\right)\right] \X \equiv
\sin\epsilon + i\cos\epsilon\,\X
\end{equation}
Then $\cos^2\epsilon$ is the reflectivity.  A mirror is simply $i\X$.

@@

\section{A classical analogy}

Consider two coupled oscillators, one of them damped.
\begin{equation}
\begin{array}{l}
\ddot x + x + \kappa y = 0 \\
\ddot y + y + \kappa x + \gamma y = 0
\end{array}
\end{equation}
If $\gamma=0$ has solutions of the form
\begin{equation}
\begin{array}{l}
x = \cos(\sqrt{1+\kappa}\,t) + \cos(\sqrt{1-\kappa}\,t)  \\
y = \cos(\sqrt{1+\kappa}\,t) - \cos(\sqrt{1-\kappa}\,t)
\end{array}
\end{equation}
and the amplitude sloshes from one to the other.

@@

\begin{figure}
\begin{center}
\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\semireflecting
\pgfsimpmirror(2,2) 
\pgfsimpmirror(2,-2)
\pgfbeamsplitter(4,0)\semireflecting
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5,-1.5)(-.7,-.7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,1.5)(3.3,.7) 
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5) 
\end{pgfpicture}
\vmargin\vmargin\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\semireflecting
\pgfsimpmirror(2,2) 
\pgfsimpmirror(2,-2)
\pgfbeamsplitter(4,0)\semireflecting
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5, 1.5)(-.7, .7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,1.5)(3.3,.7) 
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,-.7)(5.5,-1.5) 
\end{pgfpicture}
\vmargin\vmargin\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\semireflecting
\pgfabsorber(2,2) 
\pgfsimpmirror(2,-2)
\pgfbeamsplitter(4,0)\semireflecting
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5,-1.5)(-.7,-.7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5) 
\pgfxyline(4.7,-.7)(5.5,-1.5) 
\end{pgfpicture}
\end{center}
\caption{\label{fig:mz} Schematic representation of interaction-free
  measurement. Each diagram stands for a Mach-Zehnder set up so that
  both arms are exactly equal.  Accordingly, a photon sent axially
  into one arm emerges parallel to the original direction.  This is
  seen in the upper and middle diagrams.  In the lower diagram, one of
  the mirrors is replaced by an absorber, and as a result, photons can
  emerge perpendicular to their original direction.  Such photons
  effectively detect the absorber without getting absorbed.}
\end{figure}

\begin{figure}
\begin{center}
\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\nearlytransparent 
\pgfmirror(2,2) 
\pgfmirror(2,-2)
\pgfbeamsplitter(4,0)\nearlytransparent 
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5,-1.5)(-.7,-.7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,1.5)(3.3,.7) 
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5)
\pgfxyline(4.7,-.7)(5.5,-1.5)
\pgfputat{\pgfxy(-1.7,-1.7)}{\pgfbox[right,center]{$\ket0$}} 
\pgfputat{\pgfxy(5.7, 1.7)}{\pgfbox[left,center]
         {$-\sin2\epsilon\;\ket0$}}
         \pgfputat{\pgfxy(5.7,-1.7)}{\pgfbox[left,center]
         {$-i\cos2\epsilon\;\ket1$}}
\end{pgfpicture}
\vmargin\vmargin\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\nearlytransparent 
\pgfmirror(2,2) 
\pgfmirror(2,-2)
\pgfbeamsplitter(4,0)\nearlytransparent 
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5, 1.5)(-.7, .7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,1.5)(3.3,.7) 
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5)
\pgfxyline(4.7,-.7)(5.5,-1.5)
\pgfputat{\pgfxy(-1.7, 1.7)}{\pgfbox[right,center]{$\ket1$}} 
\pgfputat{\pgfxy(5.7, 1.7)}{\pgfbox[left,center]
         {$-i\cos2\epsilon\;\ket0$}}
          \pgfputat{\pgfxy(5.7,-1.7)}{\pgfbox[left,center]
         {$-\sin2\epsilon\;\ket1$}}
\end{pgfpicture}
\vmargin\vmargin\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\nearlytransparent 
\pgfabsorber(2,2) 
\pgfmirror(2,-2)
\pgfbeamsplitter(4,0)\nearlytransparent 
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5,-1.5)(-.7,-.7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5)
\pgfxyline(4.7,-.7)(5.5,-1.5)
\pgfputat{\pgfxy(-1.7,-1.7)}{\pgfbox[right,center]{$\ket0$}} 
\pgfputat{\pgfxy(5.7, 1.7)}{\pgfbox[left,center]
         {$-\sin\epsilon\cos\epsilon\;\ket0$}}
          \pgfputat{\pgfxy(5.7,-1.7)}{\pgfbox[left,center]
         {$-i\cos^2\epsilon\;\ket1$}}
\end{pgfpicture}
\end{center}
\caption{\label{fig:qubits} Like Figure \ref{fig:mz} but annotated
  with quantum states and operators.  A photon is denoted by $\ket0$
  if moving towards upper right, and by $\ket1$ if moving towards
  lower right.  Beam splitters and mirrors are unitary operators.  An
  absorber is a non-unitary operator, which zeros the $\ket0$
  subspace.}
\end{figure}

>>>>

\title{A Classical Analog of Quantum Interaction-Free Measurement}

@@

\begin{abstract}
Interaction-free measurement is usually discussed in quantum language and
refers to phenomena such as the presence of a photon absorber being
inferred without it actually absorbing any photons.  Here we make the
simple observation that there is an
exactly analogous situation in classical mechanics: If one has
a pair of identical coupled oscillators A \& B, and A is set in motion,
then one will see the strength of the oscillation of A slowly decrease
with time, and over longer times the energy will periodically `slosh' back
and forth between the two systems.  But if instead oscillator B is damped -- so
it is able to {\em absorb\/} energy -- the rate of decrease of the
energy in oscillator A will be suppressed.   
Thus one can infer the ability of oscillator B to 
absorb energy by observing the behaviour of oscillator A without, in
the limit that the damping is strong, there actually being any
energy absorbed.
% We review the physics and its realization in interferometric
% experiments.  We then consider a simple classical system: two coupled
% oscillators, one of which is damped; the energy dissipation can be
% made arbitrarily small by making the damping large enough.  The
% phenomenon can be considered analogous to interaction-free
% measurement.
\end{abstract}

@@

Many quantum paradoxes have been explored theoretically and
experimentally in recent years.  Most involve entangled states: the
famous EPR paradox is an old example, and quantum computing is a more
recent development\cite{2007qucosc.book}.  Entanglement is considered
an intrinsically quantum phenonemon with no classical analog.

There are also counter-intuitive quantum phenonema not involving
entanglement.  Imagine a photon, initially in a ``vertical'' polarized
state.  It is sent through a Faraday rotator, and then through a
vertically polarizing sheet.  The former rotates the polarization
plane towards ``horizontal'' by a small angle, say $\pi/(2N)$, where $N$
is large.  The latter effectively measures the polarization state in
the vertical/horizontal basis; if the result comes out horizontal, the
photon is absorbed, otherwise it goes through in a vertical
eigenstate.  Since $N$ is large, the chances of being absorbed are
small.  More specifically, after the rotation the {\em amplitude\/}
to be in the horizontal state will be of order $1/N$ so the probability to be
in that mode -- and therefore the probability of absorption -- is of order $1/N^2$.
Thus, most likely, the photon will get slightly rotated and then
reset to vertical.  If the photon is re-routed back to the apparatus
$N$ times, it will still most likely emerge vertically polarized.  The
rotator and the polarizer appear to cancel each other out.  The
general phenomenon is known as the quantum Zeno
effect,\cite{PhysRevA.41.2295} but it can also be described in
everyday language, in terms of waking and sleeping
states.\cite{orzel2009teach}

What happens if we now remove the polarizer?  Clearly, the $N$ Faraday
rotations will then convert the photon to a horizontal state.  Thus, the net
effect of the polarizer is to rotate a photon from
horizontal to vertical.  Despite the fact that the polarizer is really just an
absorber of horizontal photons, it can, within this experimental set-up, have
the effect of rotating photons instead of absorbing them.  This is the essence
of interaction-free measurement; we can infer, or `measure' the presence or absence
of the polarizer by measuring the polarization.  If the polarization is rotated
we know that the absorber cannot have been present -- and there is then of course
no absorption -- while if the polarization is not changed then we know that
the absorber was present, but again, at least in the limit $N \rightarrow \infty$,
no absorption actually takes place.

Since the initial suggestion
(essentially the $N=1$ case)\cite{1993FoPh...23..987E} and
generalization to the now-standard form, together with the first
experiments\cite{PhysRevLett.74.4763} interaction-free measurement has
been the subject of many
demonstrations\cite{voorthuysen:1504,2006JPhB...39.3177N} and
reviews,\cite{deweerd:272} and applications aiming towards medical
imaging are starting to appear\cite{PhysRevA.80.040902}.  (An older
usage of similar
terminology\cite{springerlink:10.1007/BF01327019,dicke:925} refers to
a situation where a non-detection enables an inference.)

@@

\section{Quantum/Wave-Mechanical formulation}

@@

With these concrete descriptions in mind, let us formalize the process
of interaction-free measurement.  We write the two eigenstates in the
notation of qbits
\begin{equation}
\ket0 \equiv \left( \begin{array}{c} 1 \\ 0 \end{array} \right)  \qquad
\ket1 \equiv \left( \begin{array}{c} 0 \\ 1 \end{array} \right)
\end{equation}
and
\begin{equation}
\X \equiv \left( \begin{array}{cc} 0 & 1 \\ 1 & 0 \end{array} \right)
\end{equation}
denotes a swap operator.  A beam splitter is given by the unitary operator
\begin{equation}
\exp\left[i\left({\textstyle\frac\pi2-\epsilon}\right)\right] \X \equiv
\sin\epsilon + i\cos\epsilon\,\X
\end{equation}
Then $\cos^2\epsilon$ is the reflectivity.  A mirror is simply $i\X$.

@@

The discussion above uses the quantum mechanical language of
`states', but the behaviour of such an interferometer can, of course,
but understood entirely as a purely classical wave obeying
Maxwell's equations.

\section{The classical analogy}

To flesh out the observation of the abstract, 
consider two coupled oscillators, one of them damped.
\begin{equation}
\begin{array}{l}
\ddot x + x + \kappa y = 0 \\
\ddot y + y + \kappa x + \gamma y = 0
\end{array}
\end{equation}
In the absence of damping ($\gamma=0$), it is easy to verify that
\begin{equation}
\begin{array}{l}
2x = \cos(\sqrt{1+\kappa}\,t) + \cos(\sqrt{1-\kappa}\,t)  \\
2y = \cos(\sqrt{1+\kappa}\,t) - \cos(\sqrt{1-\kappa}\,t)
\end{array}
\end{equation}
is a solution for the initial conditions
\begin{equation}
x(0) = 1, y(0) = \dot x(0) = \dot y(0) = 0 \,.
\end{equation}
If the coupling is weak ($\kappa\ll1$) the solution takes the form
\begin{equation}
\begin{array}{l}
x = \cos(\frac12\kappa t) \cos t \\
y = -\sin(\frac12\kappa t) \sin t \\
\end{array}
\end{equation}
and the amplitude sloshes from $x$ to $y$.  The weaker the coupling,
the slower the transfer of energy between the oscillators.

@@

\begin{figure}
\begin{center}
\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\semireflecting
\pgfsimpmirror(2,2) 
\pgfsimpmirror(2,-2)
\pgfbeamsplitter(4,0)\semireflecting
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5,-1.5)(-.7,-.7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,1.5)(3.3,.7) 
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5) 
\end{pgfpicture}
\vmargin\vmargin\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\semireflecting
\pgfsimpmirror(2,2) 
\pgfsimpmirror(2,-2)
\pgfbeamsplitter(4,0)\semireflecting
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5, 1.5)(-.7, .7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,1.5)(3.3,.7) 
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,-.7)(5.5,-1.5) 
\end{pgfpicture}
\vmargin\vmargin\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\semireflecting
\pgfabsorber(2,2) 
\pgfsimpmirror(2,-2)
\pgfbeamsplitter(4,0)\semireflecting
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5,-1.5)(-.7,-.7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5) 
\pgfxyline(4.7,-.7)(5.5,-1.5) 
\end{pgfpicture}
\end{center}
\caption{\label{fig:mz} Schematic representation of interaction-free
  measurement. Each diagram stands for a Mach-Zehnder set up so that
  both arms are exactly equal.  In the upper and middle diagrams, a
  photon sent axially into one arm emerges parallel to the original
  direction.  In the lower diagram, one of the mirrors is replaced by
  an absorber, and as a result, photons can emerge perpendicular to
  their original direction.  Such photons effectively detect the
  absorber without getting absorbed.}
\end{figure}

\begin{figure}
\begin{center}
\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\nearlytransparent 
\pgfmirror(2,2) 
\pgfmirror(2,-2)
\pgfbeamsplitter(4,0)\nearlytransparent 
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5,-1.5)(-.7,-.7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,1.5)(3.3,.7) 
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5)
\pgfxyline(4.7,-.7)(5.5,-1.5)
\pgfputat{\pgfxy(-1.7,-1.7)}{\pgfbox[right,center]{$\ket0$}} 
\pgfputat{\pgfxy(5.7, 1.7)}{\pgfbox[left,center]
         {$-\sin2\epsilon\;\ket0$}}
         \pgfputat{\pgfxy(5.7,-1.7)}{\pgfbox[left,center]
         {$-i\cos2\epsilon\;\ket1$}}
\end{pgfpicture}
\vmargin\vmargin\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\nearlytransparent 
\pgfmirror(2,2) 
\pgfmirror(2,-2)
\pgfbeamsplitter(4,0)\nearlytransparent 
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5, 1.5)(-.7, .7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,1.5)(3.3,.7) 
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5)
\pgfxyline(4.7,-.7)(5.5,-1.5)
\pgfputat{\pgfxy(-1.7, 1.7)}{\pgfbox[right,center]{$\ket1$}} 
\pgfputat{\pgfxy(5.7, 1.7)}{\pgfbox[left,center]
         {$-i\cos2\epsilon\;\ket0$}}
          \pgfputat{\pgfxy(5.7,-1.7)}{\pgfbox[left,center]
         {$-\sin2\epsilon\;\ket1$}}
\end{pgfpicture}
\vmargin\vmargin\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\nearlytransparent 
\pgfabsorber(2,2) 
\pgfmirror(2,-2)
\pgfbeamsplitter(4,0)\nearlytransparent 
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5,-1.5)(-.7,-.7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5)
\pgfxyline(4.7,-.7)(5.5,-1.5)
\pgfputat{\pgfxy(-1.7,-1.7)}{\pgfbox[right,center]{$\ket0$}} 
\pgfputat{\pgfxy(5.7, 1.7)}{\pgfbox[left,center]
         {$-\sin\epsilon\cos\epsilon\;\ket0$}}
          \pgfputat{\pgfxy(5.7,-1.7)}{\pgfbox[left,center]
         {$-i\cos^2\epsilon\;\ket1$}}
\end{pgfpicture}
\end{center}
\caption{\label{fig:qubits} Like Figure \ref{fig:mz} but annotated
  with states and operators.  A photon is denoted by $\ket0$ if moving
  towards upper right, and by $\ket1$ if moving towards lower right.
  Beam splitters and mirrors are unitary operators.  An absorber is a
  non-unitary operator, which zeros the $\ket0$ subspace.}
\end{figure}
